---
title: "Appendix: Statistical Analysis Code for DRAV Estimation"
output:
  pdf_document:
    number_sections: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# For statistical analysis
library(nlme)
library(MuMIn)
library(forecast)
library(boot)

# For plotting
library(ggplot2)
library(gridExtra)
library(ggeffects) 
library(ggpubr)
library(ggnewscale)

# For making a table
library(sjPlot)
```

# Data preparation

## Read the data

```{r load-data}
# Load and inspect the data
data <- read.csv("data/Vair_hypothesis_testing_0.03.csv")

#Criteria for analysis
data <- subset(data, avg_pitch <= -60)
data <- subset(data, glide_duration >= 10)
data <- subset(data, initial_depth <= 30)
data <- subset(data, cvar_pitch < 0.1)
data <- subset(data, cvar_roll < 0.1)
data <- subset(data, Vair > 0)
```

## Mean DRAV for each individual
```{r}
# Identify the number of DRAV estimates obtained per individual
round(tapply(data$Vair,data$individual,length),2)
# Calculate the mean and SD for the DRAV estimates for each individual
round(tapply(data$Vair,data$individual,mean),2)
round(tapply(data$Vair,data$individual,sd),2)
# Calculate the max and min for the DRAV estimates for each individual
round(tapply(data$Vair,data$individual,max),2)
round(tapply(data$Vair,data$individual,min),2)

# Find the mean and SD for the maximum dive depth
round(tapply(data$max_depth,data$individual,mean),2)
round(tapply(data$max_depth,data$individual,sd),2)
# Find the mean and SD for the average pitch during the descent glide
round(tapply(data$avg_pitch,data$individual,mean),2)
round(tapply(data$avg_pitch,data$individual,sd),2)
# Find the mean and SD for the average speed
round(tapply(data$avg_speed,data$individual,mean),2)
round(tapply(data$avg_speed,data$individual,sd),2)
# Find the mean and SD for the depth of stroke cessation
round(tapply(data$initial_depth,data$individual,mean),2)
round(tapply(data$initial_depth,data$individual,sd),2)
```

# Hypothesis testing

We will be constructing statistical models to examine the effect of body density and maximum dive depth on DRAV.

We will be assessing the following causal relationships:
- The effect of maximum dive depth on DRAV
- The effect of body density on DRAV

## Checking for auto-correlation for time-series data

```{r}
# Global Plot Options
plotopts <- theme(axis.text=element_text(size=20),axis.title = element_text(size=18),strip.text = element_text(size=20))

# Check auto-correlation of Vair:
acf(data$Vair)
```
This is the shape that you would exactly expect for an auto-correlated structure that has not been taken into account. We can try and fit a random effects model using nlme to see what auto-correlation function will take into account the auto-correlation in the model.

```{r}
# Construct the overall model with a simple AR(1) structure
lmm.full.AR1 <- lme(Vair ~ max_depth + BD + max_depth*BD,
                 random = ~1|individual, correlation = corAR1(form = ~ 1|individual),
                data = data)

summary(lmm.full.AR1)

# Extract the residuals
lmm.full.res <- residuals(lmm.full.AR1, type = "normalized")
# Check for autocorrelation:
lmm.full.acf<-autoplot(acf(lmm.full.res), main="") + theme_bw() + plotopts
```
Now auto-correlation is taken into account. This indicates that we can move on and use an AR(1) structure for further analysis.

## AIC and dredging from multimodel inferences

Using the global model that we have constructed, we will go through the dredge function to see which variable has a good model fit.
Using AIC weights, we will move onto variable selection using the dredge function. 

```{r}
### Information Theory ###
options(na.action = "na.fail") # make sure that variable selection
# will NOT undergo if there are NAs (NOT OMIT)
# This is because the number of data will make the AIC non-comparable

# Here, we will initially conduct variable selection using the AIC 
# approach; the Akaike weights of each variable

# Set "method = ML" to make sure the models are comparable.
lmm.full.test <- lme(Vair ~ max_depth + BD + max_depth*BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

# We will use the dredge function to see which model was selected.
dredge(lmm.full.test)

# Using the dredge function to compare the summed Akaike weight
# for the variables in the global model.
AIC.results <- dredge(lmm.full.test)

# The results of the Akaike weights are shown here:
sw(AIC.results)

#                     max_depth BD   BD:max_depth
# Sum of weights:      1.00 1.00      0.63        
# N containing models:    3    3         1         

# mod 1: AIC weights > 0.99
mod1 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

AIC(lmm.full.test, mod1)

# mod2 has the lowest AIC, but all models does not have a huge difference (<10). We will continue our analysis with mod1, since it is the simpler model with similar explanatory power as the full model.

# Comparing the AIC for different model structures
m1 <- lme(Vair ~ 1, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

m2 <- lme(Vair ~ max_depth, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

m3 <- lme(Vair ~ BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

m4 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

m5 <- lme(Vair ~ max_depth + BD + max_depth*BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data, method = "ML")

AIC(m1,m2,m3,m4,m5)

```

mod1, mod2 and mod3 had an AIC difference less than 10, which indicates that the two models are not too different. 

From this result, we will use mod1, the simple model for testing how the animal changes its air volume depending on dive depth and BD.

Now the final model was selected, we will look into the summary table of the final model.
```{r}
# We will now treat Individual as a random effect:
final <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                           correlation = corAR1(form = ~ 1|individual), data = data)

### Final Model ###
# The final global model:
summary(final)
summary(final)$coefficients

# Extract the correlation structure from the model
correlation_structure <- final$modelStruct$corStruct
# Extract the AR(1) coefficient (autocorrelation parameter)
ar1_coefficient <- coef(correlation_structure, unconstrained = FALSE)
ar1_coefficient

# Checking assumptions
par(mfrow=c(1,1))
# Checking the variance of the residuals
plot(final)
# Checking the distribution of the residuals
qqnorm(resid(final))
qqline(resid(final))
hist(resid(final))

```

## Prediction
Now we have selected mod1 as the model that describes how the seals vary their air volume.
Using the described model, we will calculate the predicted values of Vair with Duration and body density.

Make new data frames to predict Vair in relation to each continuous variable - this will be done by holding one of each continuous variable constant.

```{r}
## Create a dataframe for the predicted values.

# Only varying maximum dive depth
pred.df.dep <- data.frame(Vair = data$Vair, BD = mean(data$BD), max_depth = data$max_depth, individual = factor(data$individual))

# Only varying body density
pred.df.BD <- data.frame(Vair = data$Vair, BD = data$BD, max_depth = mean(data$max_depth), individual = factor(data$individual))

```

Here, we will only predict from the fixed effects from the mixed model, using the nlme package.
The fixed effect can be accessed by introducing the re.form = NA argument in the predict function - this will enable the predict function to ignore the random effect and give you the predictions based on the fixed effects.

When constructing a linear model and confidence intervals, it is best to bootstrap the values using the bootMer() function from lmer4.

```{r}
# Now getting the confidence intervals from only the fixed effects:

#-----
# Depth

# Function to apply to each bootstrap sample
boot_fn1 <- function(data, indices) {
  # Refit model on the bootstrap sample
  fit <- update(final, data = data[indices, ])
  # Predict from the refitted model
  return(predict(fit, newdata=pred.df.dep, level = 0))
}

boot.dep <- boot(data = data, statistic = boot_fn1, R = 1000)

# Calculate confidence intervals and summary statistics
pred.df.dep$Vair.pred.dep.CI.fit <- boot.dep$t0
pred.df.dep$Vair.pred.dep.CI.lwr <- apply(boot.dep$t, 2, quantile, probs = 0.025)
pred.df.dep$Vair.pred.dep.CI.upr <- apply(boot.dep$t, 2, quantile, probs = 0.975)

#-----
# BD

# Function to apply to each bootstrap sample
boot_fn3 <- function(data, indices) {
  # Refit model on the bootstrap sample
  fit <- update(final, data = data[indices, ])
  # Predict from the refitted model
  return(predict(fit, newdata=pred.df.BD, level = 0))
}

boot.BD <- boot(data = data, statistic = boot_fn3, R = 1000)

# Calculate confidence intervals and summary statistics
pred.df.BD$Vair.pred.BD.CI.fit <- boot.BD$t0
pred.df.BD$Vair.pred.BD.CI.lwr <- apply(boot.BD$t, 2, quantile, probs = 0.025)
pred.df.BD$Vair.pred.BD.CI.upr <- apply(boot.BD$t, 2, quantile, probs = 0.975)

# Insert the raw data into the dataframe (holding body density constant and duration)
pred.df.dep$Vair <- data$Vair
pred.df.dep$pred <- predict(final, data, type="response") # This is for the data points for the marginal effects plot

# Calculate predicted values (holding duration and Max_depth):
pred.df.BD$Vair <- data$Vair
pred.df.BD$pred <- predict(final, data, type="response") # This is for the data points for the marginal effects plot
```

## Data Visualisation

```{r}
# Change the name of individuals into the ID name
# Max depth
pred.df.dep$individual <- as.character(pred.df.dep$individual)

# BD
pred.df.BD$individual <- as.character(pred.df.BD$individual)

# Color-blind friendly colors:
color_palette <- c("ml17_280a" = "#009E73", "ml17_301a" = "#0072B2", "ml18_292a" = "#56B4E9", "ml18_294a" = "#F0E442", "ml18_294b" = "#E69F00", "ml19_295a" = "#D55E00")

# Plot a marginal-effects plot for Vair ~ depth
depth <- ggplot() +
  geom_point(alpha = 0.5, size = 3, data = pred.df.dep, aes(max_depth, Vair, color = "black")) + 
     geom_point(alpha = 0.5, size = 3, data = pred.df.dep, aes(max_depth, Vair, color = individual)) + 
  ylab(expression(paste("DRAV (", "ml ", kg^-1, ")", sep=""))) + 
  xlab("Maximum dive depth (m)") + 
  scale_color_manual(values = color_palette) + 
  theme_linedraw(base_size = 15) +
    theme(legend.title=element_text(size=15),
        legend.text=element_text(size=15),
        axis.title=element_text(size=15),
        axis.text=element_text(size=15))+ 
     geom_ribbon(data = pred.df.dep, aes(max_depth, Vair, ymin = Vair.pred.dep.CI.lwr, ymax = Vair.pred.dep.CI.upr, color = NULL), alpha = .2) +
     geom_line(data = pred.df.dep, aes(max_depth, Vair.pred.dep.CI.fit),  alpha = .7, size = 1) + 
  theme(legend.position = "none")+
  guides(color=guide_legend(override.aes=list(fill=NA)))

# Plot a marginal-effects plot for Vair ~ Body density
BD <- ggplot() +
  geom_point(alpha = 0.5, size = 3, data = pred.df.BD, aes(BD, Vair,  color = "black")) +
     geom_point(alpha = 0.5, size = 3, data = pred.df.BD, aes(BD, Vair,  color = individual)) + 
  ylab(expression(paste("DRAV (", "ml ", kg^-1, ")", sep=""))) +
  xlab(expression(paste(
    "Tissue density (", "kg ", m^-3, ")", sep=""))) + 
  scale_color_manual(values = color_palette) + 
  theme_linedraw(base_size = 15) +
      theme(legend.title=element_text(size=15),
        legend.text=element_text(size=15),
        axis.title=element_text(size=15),
        axis.text=element_text(size=15))+ 
  theme(legend.key=element_blank()) +
     geom_ribbon(data = pred.df.BD, aes(BD, Vair, ymin = Vair.pred.BD.CI.lwr, ymax = Vair.pred.BD.CI.upr, color = NULL), alpha = .2) +
     geom_line(data = pred.df.BD, aes(BD, Vair.pred.BD.CI.fit), 
               alpha = .7, linewidth = 1) +
  guides(color=guide_legend(override.aes=list(fill=NA)))

# Plot together
ggsave(
  "BD_MD_Vair_ggsave.png",
  ggarrange(depth, BD, ncol = 1, nrow = 2, common.legend = TRUE,
          labels=c("A", "B"),
          font.label = list(size = 25, color = "black"),
          legend="bottom"),
  width = 180,
  height = 210,
  unit = "mm",
  dpi = 1200
)
```

```{r}
# Notes: predictor labels (pred.labels) should be listed from top to bottom; dv.labels= the name of the response variable that will be at the top of the table.

sjPlot::tab_model(final,
                  pred.labels =c("(Intercept)", "Dive depth", "Body density"),
                  dv.labels= "Effects on DRAV")

```

# Depth at which Neutral Buoyancy is Achieved

```{r neutral-buoyancy-functions}
# Function to compute required air volume for neutral buoyancy
air_volume_for_NB <- function(BD, depth, g = 9.81, water_density = 1025, m = 300, Pair = 1.23) {
  pressure_term <- (water_density - Pair * (1 + 0.1 * depth)) / (1 + 0.1 * depth)
  Vair <- (m * (1 - (water_density / BD))) / pressure_term
  return(Vair)
}

# Generate NB dataframe for selected depths and BD range
create_NB_dataframe <- function(BD_range, depths = c(10, 20, 30), g = 9.81, water_density = 1025, m = 300) {
  do.call(rbind, lapply(BD_range, function(BD) {
    do.call(rbind, lapply(depths, function(depth) {
      data.frame(
        BD = BD,
        Vair = air_volume_for_NB(BD, depth, g, water_density, m),
        Depth = depth
      )
    }))
  }))
}

# Rescale utility
rescale <- function(x, from = range(x), to = c(1, 100)) {
  (x - from[1]) / diff(from) * diff(to) + to[1]
}

# Extract colour from gradient for each depth
extract_colour <- function(low, high, depths, depth_range) {
  palette <- colorRampPalette(c(low, high))(100)
  indices <- as.integer(rescale(depths, from = depth_range, to = c(1, 100)))
  palette[indices]
}
```

```{r add-NB-lines}
# Overlay theoretical NB lines with colour-matched to observed points
add_NB_lines <- function(plot, nb_df, data, m = 400) {
  nb_df$Vair_mlkg <- nb_df$Vair * 1e6 / m
  depth_range <- range(data$depth_NB)
  colours <- extract_colour("blue", "red", unique(nb_df$Depth), depth_range)

  plot +
    new_scale_color() +
    geom_line(data = nb_df, aes(x = BD, y = Vair_mlkg, colour = factor(Depth), group = Depth), size = 1.2) +
    scale_color_manual(values = setNames(colours, unique(nb_df$Depth)),
                       name = "Calculated depth at neutral buoyancy (m)") +
    theme_linedraw(base_size = 14) +
    theme(legend.position = "bottom",
          legend.title = element_text(size = 12),
          legend.text = element_text(size = 12)) +
    guides(colour = guide_legend(title.position = "top"))
}
```

```{r build-NB-plot}
# Generate BD–Vair theoretical curves
BD_range <- seq(1033.5, 1041.5, by = 0.1)
nb_df <- create_NB_dataframe(BD_range)

NB_dep <- ggplot() +
  geom_point(alpha = 0.5, size = 3, data = data, aes(BD, Vair, color = depth_NB)) + 
  ylab(expression(paste("DRAV (", "ml ", kg^-1, ")", sep=""))) +
  xlab(expression(paste(
    "Tissue density (", "kg ", m^-3, ")", sep=""))) + 
# Set a color gradient for the continuous 'BD' variable with an expression label
  scale_color_gradient(low = "blue", high = "red", 
                       name = "Depth at neutral buoyancy (m)") +
  xlim(min(data$BD) - 0.5, max(data$BD) + 0.5) + 
  ylim(min(data$Vair) - 1, max(data$Vair) + 1)

# Overlay and display
NB_dep_with_lines <- add_NB_lines(NB_dep, nb_df, data)
print(NB_dep_with_lines)

# Save figure
ggsave("NB_dep_lines.png", NB_dep_with_lines,
       width = 210, height = 160, units = "mm", dpi = 1200)
```

## Notes
- DRAV values are converted to ml/kg assuming 300 kg body mass.
- Theoretical NB curves are derived from a simplified buoyancy model.
- Colours are matched to data using a custom rescaling function.

# Sensitivity analysis to different drag coefficients

## Read in DRAV estimates from different drag coefficients

```{r}
#Criteria for analysis
criteria_Vair <- function(data) {
  data <- subset(data, avg_pitch <= -60)
  data <- subset(data, glide_duration >= 10)
  data <- subset(data, initial_depth <= 30)
  data <- subset(data, cvar_pitch < 0.1)
  data <- subset(data, cvar_roll < 0.1)
  data <- subset(data, Vair > 0)
  print(length(data$Vair))
  
  return(data)
}

df_Vair0.02 <- read.csv("data/Vair_hypothesis_testing_0.02.csv")
df_Vair0.03 <- read.csv("data/Vair_hypothesis_testing_0.03.csv")
df_Vair0.04 <- read.csv("data/Vair_hypothesis_testing_0.04.csv")
df_Vair0.05 <- read.csv("data/Vair_hypothesis_testing_0.05.csv")

df_Vair0.02 <- criteria_Vair(df_Vair0.02)
df_Vair0.03 <- criteria_Vair(df_Vair0.03)
df_Vair0.04 <- criteria_Vair(df_Vair0.04)
df_Vair0.05 <- criteria_Vair(df_Vair0.05)

```

## Find the mean and SD for the DRAV estimates for each drag coefficeint:
```{r}
round(mean(df_Vair0.02$Vair),1)
round(sd(df_Vair0.02$Vair),1)
round(mean(df_Vair0.03$Vair),1)
round(sd(df_Vair0.03$Vair),1)
round(mean(df_Vair0.04$Vair),1)
round(sd(df_Vair0.04$Vair),1)
round(mean(df_Vair0.05$Vair),1)
round(sd(df_Vair0.05$Vair),1)
```

## Find the model summary and R-squared estimates for the range of drag coefficients:
```{r}
# Checking whether the results change with different drag coefficients
final0.02 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.02)
final0.03 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.03)
final0.04 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.04)
final0.05 <- lme(Vair ~ max_depth + BD, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.05)

# Check the summary 
summary(final0.02)
summary(final0.03)
summary(final0.04)
summary(final0.05)

# Check the R-squared values
r.squaredGLMM(final0.02)
r.squaredGLMM(final0.03)
r.squaredGLMM(final0.04)
r.squaredGLMM(final0.05)
```

## Find the model summary and R-squared estimates for the kinematic correlation:

```{r}
#--
# Depth of stroke cessation ~ DRAV
m0.02 <- lme(initial_depth ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.02)
m0.03 <- lme(initial_depth ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.03)
m0.04 <- lme(initial_depth ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.04)
m0.05 <- lme(initial_depth ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_Vair0.05)

# Check the summary 
summary(m0.02)
summary(m0.03)
summary(m0.04)
summary(m0.05)

# Check the R-squared values
r.squaredGLMM(m0.02)
r.squaredGLMM(m0.03)
r.squaredGLMM(m0.04)
r.squaredGLMM(m0.05)

#--
# RMS sway dynamic acceleration ~ DRAV
# Reading the Vair estimates with the kinematic parameters
df_stroke0.02 <- read.csv("data/stroke_Vair_analysis_0.02.csv")
df_stroke0.03 <- read.csv("data/stroke_Vair_analysis_0.03.csv")
df_stroke0.04 <- read.csv("data/stroke_Vair_analysis_0.04.csv")
df_stroke0.05 <- read.csv("data/stroke_Vair_analysis_0.05.csv")

df_stroke0.02 <- criteria_Vair(df_stroke0.02)
df_stroke0.03 <- criteria_Vair(df_stroke0.03)
df_stroke0.04 <- criteria_Vair(df_stroke0.04)
df_stroke0.05 <- criteria_Vair(df_stroke0.05)

m0.02 <- lme(total_rms ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_stroke0.02)
m0.03 <- lme(total_rms ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_stroke0.03)
m0.04 <- lme(total_rms ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_stroke0.04)
m0.05 <- lme(total_rms ~ Vair, random = ~1|individual, 
                 correlation = corAR1(form = ~ 1|individual), data = df_stroke0.05)

# Check the summary 
summary(m0.02)$tTable
summary(m0.03)$tTable
summary(m0.04)$tTable
summary(m0.05)$tTable

# Check the R-squared values
r.squaredGLMM(m0.02)
r.squaredGLMM(m0.03)
r.squaredGLMM(m0.04)
r.squaredGLMM(m0.05)

```

